---
title: "PM566: Data Analysis for Covid 19 Vaccine Hesitancy and possible demographic and geographic correlations"
author: "Nicole Tang"
format: html
embed-resources: true
fig-width: 9
fig-height: 6
---
```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(leaflet)
library(tidyverse)
library(data.table)
library(R.utils)
library(dplyr)
library(ggplot2)
library(lubridate)
library(gganimate)
library(ggforce)
library(tidytext)
library(tidyr)
library(readr)
library(dplyr)
library(knitr)
library(kableExtra)
library(plotly)
```
## Description


[View the Full Analysis in My Written Report](https://github.com/nktang05/PM566Final/blob/main/Final%20Project.pdf)


***Introduction***

COVID-19 vaccine hesitancy refers to the reluctance or refusal to get vaccinated despite the availability of vaccines. Vaccination plays a crucial role in controlling the pandemic by reducing the spread of the virus, preventing severe illness, and decreasing hospitalization and death rates. However, hesitancy has been influenced by factors such as misinformation, distrust in healthcare systems or government authorities, concerns about the speed of vaccine development, and fears about potential side effects. Social, cultural, and political contexts have also shaped people’s attitudes toward vaccines. 

This data set has various demographic information showing information by county, state, ethnicity, COVID-19 vaccine coverage (CVAC) and social vulnerability index (SVI). In order to determine hesitancy levels, people were surveyed  “Once a vaccine to prevent COVID-19 is available to you, would you…get a vaccine?” and the following options were: 1) “definitely get a vaccine”; 2) “probably get a vaccine”; 3) “unsure”; 4) “probably not get a vaccine”; 5) “definitely not get a vaccine”. his data set also looks into varying levels of hesitancy: hesitant, hesitant or unsure, or strongly hesitant. People who responded “probably not” or “definitely not” were categorized as hesitant.

Data set origin: https://data.cdc.gov/Vaccinations/Vaccine-Hesitancy-for-COVID-19-County-and-local-es/q9mh-h2tw/about_data


**Research Questions**

Are there any correlations between state and region and the rates of vaccine hesitancy?
Are there any correlations between the social vulnerability index and rates of vaccine hesitancy?
Are there any correlations between COVID-19 vaccine coverage and rates of vaccine hesitancy?
Are there any correlations between ethnicity and rates of vaccine hesitancy?


## Analysis
```{r, echo = FALSE}
#read in csv
data <- read_csv("Vaccine_Hesitancy_for_COVID-19__County_and_local_estimates_20241025.csv", show_col_types = FALSE)

#test for success
#head(data)

initial_row_count <- nrow(data)

#remove rows with NA
data <- data %>%
  drop_na()

final_row_count <- nrow(data)

rows_dropped <- initial_row_count - final_row_count
#print(rows_dropped)

#mutate data type of long and lat
data <- data %>%
  mutate(
    geo_point_clean = str_remove_all(`Geographical Point`, "POINT \\(|\\)"),
    long = as.numeric(str_split_fixed(geo_point_clean, " ", 2)[, 1]),
    lat = as.numeric(str_split_fixed(geo_point_clean, " ", 2)[, 2])
  ) %>%
  select(-geo_point_clean)

region_lookup <- data.frame(
  State = c("CONNECTICUT", "MAINE", "MASSACHUSETTS", "NEW HAMPSHIRE", "RHODE ISLAND", "VERMONT","DISTRICT OF COLUMBIA", 
            "NEW JERSEY", "NEW YORK", "PENNSYLVANIA", 
            "ILLINOIS", "INDIANA", "IOWA", "KANSAS", "MICHIGAN", "MINNESOTA", "MISSOURI", "NEBRASKA", "NORTH DAKOTA", 
            "OHIO", "SOUTH DAKOTA", "WISCONSIN", 
            "ALABAMA", "ARKANSAS", "DELAWARE", "FLORIDA", "GEORGIA", "KENTUCKY", "LOUISIANA", "MARYLAND", "MISSISSIPPI", 
            "NORTH CAROLINA", "OKLAHOMA", "SOUTH CAROLINA", "TENNESSEE", "TEXAS", "VIRGINIA", "WEST VIRGINIA", 
            "ALASKA", "ARIZONA", "CALIFORNIA", "COLORADO", "HAWAII", "IDAHO", "MONTANA", "NEVADA", "NEW MEXICO", 
            "OREGON", "UTAH", "WASHINGTON", "WYOMING"),
  Region = c(rep("Northeast", 10), rep("Midwest", 12), rep("South", 17), rep("West", 12))
)

data <- data %>%
  left_join(region_lookup, by = "State")

#test for success
#summary(data)

```

```{r, echo = FALSE}

#calculate the average estimated hesitant rate per state and add to new df
mean_hesitancy <- aggregate(data$'Estimated hesitant', by = list(data$'State Code'), FUN = mean, na.rm = TRUE)
colnames(mean_hesitancy) <- c("State_Code", "Estimated_Hesitant")

#calculate not hesitant rate
data$Estimated_not_hesitant <- 100 - (data$'Estimated hesitant' + 
                                        data$'Estimated hesitant or unsure' + 
                                        data$'Estimated strongly hesitant')

#calculate the not hesitant rate per state and add to new df
mean_not_hesitant <- aggregate(data$'Estimated_not_hesitant', by = list(data$'State Code'), FUN = mean, na.rm = TRUE)
colnames(mean_not_hesitant) <- c("State_Code", "Estimated_not_hesitant")

#join the two dfs using State
mean_hesitancy <- merge(mean_hesitancy, mean_not_hesitant, by = "State_Code")

#make another df with lat and long
state_coords <- unique(data[, c("State Code", "long", "lat")])

# Merge long and lat df
mean_hesitancy <- merge(mean_hesitancy, state_coords, by.x = "State_Code", by.y = "State Code", all.x = TRUE)

region_lookup <- data.frame(
  State_Code = c("CT", "ME", "MA", "NH", "RI", "VT", "NJ", "NY", "PA", "DC",
                   "IL", "IN", "IA", "KS", "MI", "MN", "MO", "NE", "ND", "OH", "SD", "WI",
                   "AL", "AR", "DE", "FL", "GA", "KY", "LA", "MD", "MS", "NC", "OK", "SC", "TN", "TX", "VA", "WV",
                   "AK", "AZ", "CA", "CO", "HI", "ID", "MT", "NV", "NM", "OR", "UT", "WA", "WY"),
  Region = c(rep("Northeast", 10), rep("Midwest", 12), rep("South", 17), rep("West", 12))
)

mean_hesitancy <- mean_hesitancy %>%
  left_join(region_lookup, by = "State_Code")


#check for df successful merging
#head(mean_hesitancy)
```

```{r, echo = FALSE}
summary_table <- mean_hesitancy %>%
  summarise(
    Mean_Hesitant = mean(Estimated_Hesitant, na.rm = TRUE),
    SD_Hesitant = sd(Estimated_Hesitant, na.rm = TRUE),
    Min_Hesitant = min(Estimated_Hesitant, na.rm = TRUE),
    Max_Hesitant = max(Estimated_Hesitant, na.rm = TRUE),
    Mean_Not_Hesitant = mean(Estimated_not_hesitant, na.rm = TRUE),
    SD_Not_Hesitant = sd(Estimated_not_hesitant, na.rm = TRUE),
    Min_Not_Hesitant = min(Estimated_not_hesitant, na.rm = TRUE),
    Max_Not_Hesitant = max(Estimated_not_hesitant, na.rm = TRUE)
  )

# Use kable to create a publication-quality table
table <- kable(summary_table, caption = "Summary Statistics for Vaccine Hesitancy",
      col.names = c("Mean Hesitant (%)", "SD Hesitant", "Min Hesitant (%)", "Max Hesitant (%)",
                    "Mean Not Hesitant (%)", "SD Not Hesitant", "Min Not Hesitant (%)", "Max Not Hesitant (%)")) %>%
  kable_styling(full_width = FALSE, position = "center")

#table
```


```{r, echo = FALSE}
ethnicities <- c("Percent Hispanic", 
                 "Percent non-Hispanic American Indian/Alaska Native",
                 "Percent non-Hispanic Asian",
                 "Percent non-Hispanic Black",
                 "Percent non-Hispanic Native Hawaiian/Pacific Islander",
                 "Percent non-Hispanic White")

# Use max.col to determine the predominant ethnicity
data$Predominant_Ethnicity <- colnames(data[ethnicities])[max.col(data[ethnicities], ties.method = "first")]

# Print the first few rows to verify the new column
#head(data[c("State", "Predominant_Ethnicity")])

# Calculate the average estimated hesitant percentage by ethnicity
ethnicity_df <- aggregate(data$'Estimated hesitant', by = list(data$'Predominant_Ethnicity'), FUN = mean, na.rm = TRUE)
colnames(ethnicity_df) <- c("ethnicity", "Estimated_Hesitant")

ethnicity_df$ethnicity <- gsub("Percent ", "", ethnicity_df$ethnicity)
```

***Visualization Results***




```{r, echo = FALSE}
# Order descending by mean hesitancy
mean_hesitancy <- mean_hesitancy %>%
  arrange(desc(Estimated_Hesitant))

# Ensure State_Code is a factor and ordered by Estimated_Hesitant
mean_hesitancy$State_Code <- factor(mean_hesitancy$State_Code, levels = mean_hesitancy$State_Code)

# Ensure Region is also a factor
mean_hesitancy$Region <- factor(mean_hesitancy$Region)

# Create the bar chart
v1 <- ggplot(mean_hesitancy, aes(x = State_Code, y = Estimated_Hesitant, fill = Region, text = paste("Region:", Region, "<br>State:", State_Code, "<br>Hesitancy Rate:", round(Estimated_Hesitant, 2)))) +
      geom_bar(stat = "identity") +
      labs(
        title = "Fig 1. Average COVID-19 Vaccine Hesitancy by State",
        x = "State",
        y = "Average Estimated Hesitant (%)"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_text(size = 8, angle = 45, hjust = 1), 
        axis.text.y = element_text(size = 8)
      ) +
      scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
      scale_fill_brewer(palette = "Set1")

# Convert the ggplot to an interactive plotly plot
v1_interactive <- ggplotly(v1, tooltip = "text")

# Display the interactive plot
v1_interactive

```
**Figure 1: Average Hesitancy Rates by State**
Montana has the highest rates of hesitancy at 25.14%. It is followed by Wyoming (21.78%) and Arkansas (21.77%) who have similar rates of hesitancy. Vermont has the lowest rates of hesitancy (4.03%). Based on this visualization and the difference from maximum to minimum of estimate rates I would conclude that hesitancy rates and states are correlated. The color coded grouping of region also allows to see if there is a possible correlation between region and vaccine hesitancy. Based on the figure the South appears to be more hesitant and the Northeast on average is less hesitant.

```{r, echo = FALSE}
#order by hesitancy rates
mean_hesitancy <- mean_hesitancy %>%
  arrange(desc(Estimated_Hesitant))

#make sure factor
mean_hesitancy$State_Code <- factor(mean_hesitancy$State_Code, levels = mean_hesitancy$State_Code)

#new df to look at hesitant and not hesitant rates
mean_hesitancy_long <- mean_hesitancy %>%
  pivot_longer(cols = c("Estimated_Hesitant", "Estimated_not_hesitant"), 
               names_to = "Type", 
               values_to = "Percentage")


# Create a scatter plot
v2.5 <- ggplot(mean_hesitancy, aes(x = Estimated_Hesitant, y = Estimated_not_hesitant, label = State_Code)) +
  geom_point(aes(color = Region), size = 4, alpha = 0.8) +
  geom_text(vjust = -0.5, size = 3) +
  labs(
    title = "Fig 2. Scatter Plot of Hesitancy vs. Non-Hesitancy by State",
    x = "Average Estimated Hesitant (%)",
    y = "Average Estimated Not Hesitant (%)",
    color = "Region"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, hjust = 0.5)
  ) +
  scale_color_brewer(palette = "Set1")

# Convert the ggplot to an interactive plotly plot
v2.5_interactive <- ggplotly(v2.5, tooltip = c("x", "y", "label", "color"))

# Display the interactive plot
v2.5_interactive

```
**Figure 2: Average Hesitancy Rates and Non Hesitancy Rates by State**
Montana has the highest rates of hesitancy (25.14%) and the lowest rates of non hesitancy (27.74%). It is followed by Wyoming (21.78%) and Arkansas (21.77%) who have similar rates of hesitancy. However, Arkansas (36.05%) has higher rates of non hesitancy than Wyoming (33.96%). Vermont (4.03%) has the lowest rates of hesitancy and Massachusetts (4.63%) has the second lowest rates of hesitancy. However Massachusetts (85.13%) has the highest rates of non hesitancy. Based on this visualization I would conclude that hesitancy and non hesitancy appear to be inversely related and the higher the rate of hesitancy, the lower the rate of non hesitancy. I would also conclude that the variance of mean hesitancy rates and non hesitancy rates differ by states indication a correlation between state and hesitancy rates. By region, the south appears to be clustered together with higher hesitancy rates and lower not hesitant rates. The Northeast appears to be clustered with low rates of hesitancy and high rates on non hesitancy.

```{r, echo = FALSE}
data$`SVI Category` <- factor(data$`SVI Category`, levels = c("Very Low Vulnerability", 
                                                              "Low Vulnerability", 
                                                              "Moderate Vulnerability", 
                                                              "High Vulnerability", 
                                                              "Very High Vulnerability"))

# Create the box plot with ordered categories

# Create the box plot using ggplot
v4 <- ggplot(data, aes(x = `SVI Category`, y = `Estimated hesitant`, fill = `SVI Category`, text = paste(
  "SVI Category:", `SVI Category`, "<br>",
  "Estimated Hesitancy:", round(`Estimated hesitant`, 2)
))) +
  geom_boxplot() +
  labs(
    title = "Fig 3. Box Plot of COVID-19 Vaccine Hesitancy by Social Vulnerability",
    x = "Social Vulnerability Category",
    y = "Estimated Hesitancy (%)"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12, face = "bold"), 
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    legend.position = "none"  # Remove the legend
  )

# Convert the ggplot to an interactive plotly plot
v4_interactive <- ggplotly(v4, tooltip = "text")

# Display the interactive plot
v4_interactive
```
**Figure 3: Average Hesitancy Rates by Social Vulnerability Index**
Very high vulnerability has the highest median estimated hesitancy (16.76%). Very low vulnerability has the lowest median rates of hesitancy (10.55%). This is interesting because you would think that the higher vulnerability would not be quite so hesitant.

```{r, echo = FALSE}
data$`CVAC Level Of Concern` <- factor(data$`CVAC Level Of Concern`, levels = c("Very Low Concern", 
                                                              "Low Concern", 
                                                              "Moderate Concern", 
                                                              "High Concern", 
                                                              "Very High Concern"))

# Create the box plot with ordered categories

# Create the box plot using ggplot
v4.5 <- ggplot(data, aes(x = `CVAC Level Of Concern`, y = `Estimated hesitant`, fill = `CVAC Level Of Concern`, text = paste(
  "CVAC Level Of Concern:", `CVAC Level Of Concern`, "<br>",
  "Estimated Hesitancy:", round(`Estimated hesitant`, 2)
))) +
  geom_boxplot() +
  labs(
    title = "Fig 4. Box Plot of COVID-19 Vaccine Hesitancy by CVAC Level Of Concern",
    x = "CVAC Level Of Concern",
    y = "Estimated Hesitancy (%)"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12, face = "bold"), 
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    legend.position = "none"  # Remove the legend
  )

# Convert the ggplot to an interactive plotly plot
v4.5_interactive <- ggplotly(v4.5, tooltip = "text")

# Display the interactive plot
v4.5_interactive
```
**Figure 4: Average Hesitancy Rates by CVAC**
Very high concern has the highest median estimated hesitancy (16.80%). Low concern has the lowest median rates of hesitancy (10.18%). This makes sense that areas where there is very high concern of vaccine rollout challenges could be high levels of hesitancy. For example, misinformation could be the cause of high levels of hesitancy and cause challenges to vaccine rollouts.

```{r, echo = FALSE}
data_eth <- data %>%
  pivot_longer(
    cols = c("Percent Hispanic", 
             "Percent non-Hispanic American Indian/Alaska Native",
             "Percent non-Hispanic Asian",
             "Percent non-Hispanic Black",
             "Percent non-Hispanic Native Hawaiian/Pacific Islander",
             "Percent non-Hispanic White"),
    names_to = "Ethnicity",
    values_to = "Ethnicity_Percentage"
  )

v6 <- ggplot(data_eth, aes(x = Ethnicity_Percentage, y = `Estimated hesitant`, color = Ethnicity, text = paste(
  "Ethnicity: ", Ethnicity, "<br>",
  "Ethnicity Percentage: ", round(Ethnicity_Percentage, 2), "%<br>",
  "Estimated Hesitant: ", round(`Estimated hesitant`, 2), "%"
))) +
  geom_point(alpha = 0.7) + 
  labs(
    title = "Fig 5.Scatter Plot of Vaccine Hesitancy vs. Ethnicity Percentage",
    x = "Ethnicity Percentage (%)",
    y = "Estimated Hesitancy (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.position = "none"
  ) +
  facet_wrap(~ Ethnicity, scales = "free_x")

# Convert to an interactive plotly plot
v6_interactive <- ggplotly(v6, tooltip = "text")

# Display the interactive plot
v6_interactive
```
**Figure 5: Average Hesitancy Rates by Ethnicity**
Some groups like non-Hispanic Asians appear to have lower overall hesitancy, while groups such as non-Hispanic Black and non-Hispanic American Indian/Alaska Native show a wider spread and higher average hesitancy.








